---
title: "What is Statistical Learning?"
subtitle: "Chapter 02 – Part I"
date: today
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    chalkboard: true
    #multiplex: true
    preview-links: auto
    logo: ""
    footer: "ML in Economics | [Zahid Asghar](https://www.zahid.quarto.pub)"
    transition: slide
    background-transition: fade
    highlight-style: github
    code-fold: true
    code-tools: true
    menu:
      side: left
      width: normal
    css: custom.scss
execute:
  echo: false
  warning: false
  freeze: auto 
  
---

## Outline
* What is Statistical Learning?
* Why estimate $f$?
* How do we estimate $f$?
* The trade-off: prediction accuracy vs. model interpretability
* Supervised vs. unsupervised learning
* Regression vs. classification

---

## What is Statistical Learning?

Think of statistical learning like **learning to predict** based on patterns in data.

**Simple Example**: Predicting house prices based on size, location, age.

Data: $(X_1, X_2, \dots, X_p, Y)$ for $i = 1, \dots, n$. 
  - $X_j$ are predictors (size, location, age of house)
  - $Y$ is the response (house price)

**Key Assumption**: House price depends on at least one feature (size matters!)

**Model**: $Y = f(X) + \varepsilon$ 
- $f$ = the "true relationship" we want to learn
- $\varepsilon$ = random noise (things we can't predict)

> Statistical learning finds the best $f$ using data

---

## Why Do We Want to Estimate $f$?

**Two main reasons:**

::: {.incremental}
1. **Prediction** 📈
   - "What will this house sell for?"
   - "Will this customer buy our product?"
   - We don't care HOW the prediction works, just that it's accurate

2. **Understanding** 🔍  
   - "How does education affect wages?"
   - "Which marketing channels work best?"
   - We want to understand the relationship itself
:::

**Real-world example**: A bank wants to predict loan defaults (prediction) AND understand what factors cause defaults (understanding)

---

## Example: Wage Data

::: {.panel-tabset}

## R Code
```{r}
#| echo: true
#| eval: false
library(ISLR2)
library(ggplot2)
library(dplyr)

data(Wage)
Wage %>%
  ggplot(aes(x = age, y = wage)) +
  geom_point(alpha = 0.4, color = "steelblue") +
  geom_smooth(method = "loess", se = FALSE, color = "red") +
  labs(title = "Wage vs. Age", x = "Age", y = "Wage ($1000s)") +
  theme_minimal()
```

## Python Code
```python
import pandas as pd
import matplotlib.pyplot as plt
import seaborn as sns
from sklearn.datasets import fetch_openml

# Load wage data (or similar dataset)
wage_data = pd.read_csv('wage_data.csv')

plt.figure(figsize=(10, 6))
sns.scatterplot(data=wage_data, x='age', y='wage', alpha=0.4)
sns.loess(data=wage_data, x='age', y='wage', color='red')
plt.title('Wage vs. Age')
plt.xlabel('Age')
plt.ylabel('Wage ($1000s)')
plt.show()
```

## Interpretation
**What we see**: 
- Wages increase with age until around 40-50
- Then wages level off or slightly decrease
- Lots of variation at each age (the "noise")

**Business insight**: Companies should expect to pay higher wages to experienced workers, but the premium levels off
:::

```{r}
#| echo: false
library(ISLR2)
library(ggplot2)
library(dplyr)

data(Wage)

Wage %>%
  ggplot(aes(x = age, y = wage)) +
  geom_point(alpha = 0.4, color = "steelblue") +
  geom_smooth(method = "loess", se = FALSE, color = "red", linewidth = 1.2) +
  labs(
    title = "Wage vs. Age with LOESS Fit",
    x = "Age",
    y = "Wage ($1000s)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 11)
  )
```

---

## How Noise Affects Learning

**Key Insight**: More noise = harder to learn the pattern

Think of it like listening to music with background static - more static makes it harder to hear the song.

::: {.panel-tabset}

## R Code
```{r}
#| echo: true
#| eval: false
set.seed(123)
x <- seq(-2, 2, length.out = 100)
y_true <- sin(pi * x)  # True relationship
y_low <- y_true + rnorm(100, 0, 0.2)   # Low noise
y_high <- y_true + rnorm(100, 0, 1)    # High noise

# Create comparison plots
library(patchwork)
p1 <- ggplot(data.frame(x=x, y=y_low), aes(x, y)) + 
      geom_point(color="steelblue") + 
      geom_line(aes(y=y_true), color="red") +
      ggtitle("Low Noise - Easy to Learn")

p2 <- ggplot(data.frame(x=x, y=y_high), aes(x, y)) + 
      geom_point(color="steelblue") + 
      geom_line(aes(y=y_true), color="red") +
      ggtitle("High Noise - Hard to Learn")

p1 + p2
```

## Python Code
```python
import numpy as np
import matplotlib.pyplot as plt

np.random.seed(123)
x = np.linspace(-2, 2, 100)
y_true = np.sin(np.pi * x)
y_low = y_true + np.random.normal(0, 0.2, 100)
y_high = y_true + np.random.normal(0, 1, 100)

fig, (ax1, ax2) = plt.subplots(1, 2, figsize=(12, 5))

ax1.scatter(x, y_low, alpha=0.7, color='steelblue')
ax1.plot(x, y_true, color='red', linewidth=2)
ax1.set_title('Low Noise - Easy to Learn')

ax2.scatter(x, y_high, alpha=0.7, color='steelblue')  
ax2.plot(x, y_true, color='red', linewidth=2)
ax2.set_title('High Noise - Hard to Learn')

plt.tight_layout()
plt.show()
```

## Business Application
**Real Example**: Predicting sales

- **Low noise**: Luxury cars (few, predictable buyers)
- **High noise**: Fast food (many random factors)

**Strategy**: With high noise, collect MORE data or use simpler models
:::

```{r}
#| echo: false
library(patchwork)

set.seed(123)
x <- seq(-2, 2, length.out = 100)
y_true <- sin(pi * x)
y_low <- y_true + rnorm(100, 0, 0.2)
y_high <- y_true + rnorm(100, 0, 1)

df_low <- data.frame(x = x, y = y_low, y_true = y_true)
df_high <- data.frame(x = x, y = y_high, y_true = y_true)

p1 <- df_low %>%
  ggplot(aes(x = x)) +
  geom_point(aes(y = y), color = "steelblue", alpha = 0.7, size = 1.5) +
  geom_line(aes(y = y_true), color = "red", linewidth = 1.2) +
  labs(title = "Low Noise - Easy to Learn", x = "x", y = "y") +
  theme_minimal()

p2 <- df_high %>%
  ggplot(aes(x = x)) +
  geom_point(aes(y = y), color = "steelblue", alpha = 0.7, size = 1.5) +
  geom_line(aes(y = y_true), color = "red", linewidth = 1.2) +
  labs(title = "High Noise - Hard to Learn", x = "x", y = "y") +
  theme_minimal()

p1 + p2
```

---

## Different Ways to Estimate $f$

**Question**: Given data points, how do we draw the "best" line?

**Answer**: Different methods give different results!

::: {.panel-tabset}

## R Code  
```{r}
#| echo: true
#| eval: false
# Method 1: Linear regression (straight line)
linear_fit <- lm(y ~ x, data = df)

# Method 2: Polynomial (curved line) 
poly_fit <- lm(y ~ poly(x, 3), data = df)

# Method 3: LOESS (flexible curve)
df %>%
  ggplot(aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", color = "red") +      # Linear
  geom_smooth(method = "loess", color = "blue") +  # Flexible
  labs(title = "Different Methods, Different Results")
```

## Python Code
```python
from sklearn.linear_model import LinearRegression
from sklearn.preprocessing import PolynomialFeatures
from sklearn.pipeline import Pipeline
import numpy as np

# Linear regression
linear = LinearRegression()
linear.fit(x.reshape(-1,1), y)

# Polynomial regression  
poly = Pipeline([
    ('poly', PolynomialFeatures(degree=3)),
    ('linear', LinearRegression())
])
poly.fit(x.reshape(-1,1), y)

# Plot results
plt.scatter(x, y)
plt.plot(x, linear.predict(x.reshape(-1,1)), 'r-', label='Linear')
plt.plot(x, poly.predict(x.reshape(-1,1)), 'b-', label='Polynomial')
plt.legend()
plt.show()
```

## Key Tradeoff
**Simple models** (straight lines):
- ✅ Easy to understand and explain
- ❌ Might miss important patterns

**Complex models** (wiggly curves):
- ✅ Capture complex patterns  
- ❌ Hard to interpret, might overfit

**Rule of thumb**: Start simple, add complexity only if needed
:::

```{r}
#| echo: false
df_smooth <- data.frame(x = x, y = y_low, y_true = y_true)

df_smooth %>%
  ggplot(aes(x = x)) +
  geom_point(aes(y = y), color = "grey60", alpha = 0.8, size = 1.5) +
  geom_line(aes(y = y_true), color = "red", linewidth = 1.5, linetype = "solid") +
  geom_smooth(aes(y = y), method = "loess", se = FALSE, color = "blue", linewidth = 1.2, span = 0.3) +
  labs(
    title = "Function Estimation with LOESS",
    subtitle = "Red = True relationship, Blue = Our estimate",
    x = "x", 
    y = "y"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 13, face = "bold")
  )
```

---

## Real Example: Boston Housing Prices

**Business Question**: How much should we price this house?

::: {.panel-tabset}

## R Code
```{r}
#| echo: true
#| eval: false
library(MASS)
data(Boston)

# Simple model: Price depends on poverty level and room count
lm_fit <- Boston %>%
  lm(medv ~ lstat + rm, data = .)

# Look at results
summary(lm_fit)

# Interpretation:
# - Each extra room adds ~$9,000 to house value
# - 1% increase in poverty rate reduces value by ~$950
```

## Python Code  
```python
import pandas as pd
from sklearn.linear_model import LinearRegression
from sklearn.datasets import load_boston

# Load Boston housing data
boston = load_boston()
df = pd.DataFrame(boston.data, columns=boston.feature_names)
df['price'] = boston.target

# Fit model: Price = f(poverty_rate, avg_rooms)
model = LinearRegression()
X = df[['LSTAT', 'RM']]  # poverty %, avg rooms
y = df['price']

model.fit(X, y)

# Print coefficients
print(f"Room effect: ${model.coef_[1]:.2f}k per room")
print(f"Poverty effect: -${abs(model.coef_[0]):.2f}k per 1% increase")
```

## Business Insights
**Actionable findings**:

1. **Room count matters**: Each extra room = ~$9K more value
   - Strategy: Highlight room count in listings
   
2. **Location matters**: High-poverty areas have lower prices  
   - Strategy: Different pricing strategies by neighborhood
   
3. **Model fit**: R² ≈ 0.67 means we explain 67% of price variation
   - Still missing some important factors (schools, transportation?)
:::

```{r}
#| echo: false
library(MASS)
library(broom)
data(Boston)

lm_fit <- Boston %>%
  lm(medv ~ lstat + rm, data = .)

# Create a nice summary table
tidy_results <- tidy(lm_fit) %>%
  mutate(
    term = case_when(
      term == "(Intercept)" ~ "Baseline Price",
      term == "lstat" ~ "Poverty Rate Effect (%)", 
      term == "rm" ~ "Additional Room Effect"
    ),
    estimate = round(estimate, 2),
    p.value = case_when(
      p.value < 0.001 ~ "< 0.001",
      TRUE ~ as.character(round(p.value, 3))
    )
  )

knitr::kable(tidy_results, 
             col.names = c("Factor", "Effect ($1000s)", "Std Error", "t-stat", "p-value"),
             caption = "What Affects Boston Housing Prices?")
```

---

## Visualizing Housing Price Relationships

**Key insight**: Relationships aren't always linear!

```{r}
#| echo: false
library(patchwork)

p1 <- Boston %>%
  ggplot(aes(x = lstat, y = medv)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_smooth(method = "lm", se = FALSE, color = "red", linewidth = 1.2) +
  geom_smooth(method = "loess", se = FALSE, color = "blue", linewidth = 1.2, linetype = "dashed") +
  labs(
    title = "Poverty Rate vs. Price",
    subtitle = "Red = Linear fit, Blue = Flexible fit", 
    x = "% Lower Status Population",
    y = "Median Home Value ($1000s)"
  ) +
  theme_minimal()

p2 <- Boston %>%
  ggplot(aes(x = rm, y = medv)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_smooth(method = "lm", se = FALSE, color = "red", linewidth = 1.2) +
  geom_smooth(method = "loess", se = FALSE, color = "blue", linewidth = 1.2, linetype = "dashed") +
  labs(
    title = "Room Count vs. Price",
    subtitle = "Red = Linear fit, Blue = Flexible fit",
    x = "Average Number of Rooms",
    y = "Median Home Value ($1000s)"
  ) +
  theme_minimal()

p1 + p2
```

**Business Implication**: The poverty-price relationship is clearly non-linear! Linear models miss this pattern.

---

## Supervised vs. Unsupervised Learning {.scrollable}

::: {.columns}

::: {.column width="50%"}
### Supervised 👨‍🏫
**You have the "answer key"**

Examples:
- Predict house prices (have past sales)
- Classify emails as spam (have labeled examples)
- Forecast sales (have historical data)

**Goal**: Learn from examples to predict new cases
:::

::: {.column width="50%"}
### Unsupervised 🔍  
**No "answer key" - find hidden patterns**

Examples:
- Group customers by behavior
- Find market segments
- Detect unusual transactions

**Goal**: Discover structure in data
:::

:::

::: {.panel-tabset}

## R Code
```{r}
#| echo: true  
#| eval: false
# Unsupervised: Find customer segments
set.seed(2)
customer_data <- tibble(
  spending = c(rnorm(50, 30, 10), rnorm(50, 80, 15)),
  frequency = c(rnorm(50, 2, 1), rnorm(50, 8, 2))
)

# K-means clustering (finds groups automatically)
segments <- kmeans(customer_data, centers = 2)
customer_data$segment <- segments$cluster

# Plot results  
ggplot(customer_data, aes(spending, frequency, color = factor(segment))) +
  geom_point(size = 3) +
  labs(title = "Customer Segments Found Automatically")
```

## Python Code
```python
from sklearn.cluster import KMeans
import matplotlib.pyplot as plt

# Generate customer data
spending = np.concatenate([np.random.normal(30, 10, 50), 
                          np.random.normal(80, 15, 50)])
frequency = np.concatenate([np.random.normal(2, 1, 50),
                           np.random.normal(8, 2, 50)])

# Find segments
kmeans = KMeans(n_clusters=2)
segments = kmeans.fit_predict(np.column_stack([spending, frequency]))

# Plot
plt.scatter(spending, frequency, c=segments, cmap='viridis')
plt.xlabel('Average Spending')
plt.ylabel('Purchase Frequency') 
plt.title('Customer Segments Found Automatically')
plt.show()
```

## Business Value
**Unsupervised learning reveals**:

1. **Customer segments** → Targeted marketing
2. **Fraud patterns** → Risk management  
3. **Market structure** → Competitive strategy

**Key insight**: Sometimes the most valuable discoveries come from data exploration, not prediction!
:::

```{r}
#| echo: false
set.seed(2)

cluster_data <- bind_rows(
  tibble(spending = rnorm(50, 30, 10), frequency = rnorm(50, 2, 1)),
  tibble(spending = rnorm(50, 80, 15), frequency = rnorm(50, 8, 2))
)

km <- kmeans(cluster_data, centers = 2, nstart = 20)
cluster_data$segment <- factor(km$cluster, labels = c("Low Value", "High Value"))

cluster_data %>%
  ggplot(aes(x = spending, y = frequency, color = segment)) +
  geom_point(size = 2.5, alpha = 0.8) +
  scale_color_manual(values = c("red", "blue")) +
  labs(
    title = "Customer Segments Discovered Automatically",
    subtitle = "No prior labels needed - algorithm found the groups",
    x = "Average Spending ($)", 
    y = "Purchase Frequency",
    color = "Segment"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

---

## Regression vs. Classification {.scrollable}

**The key difference**: What type of outcome are you predicting?

::: {.columns}
::: {.column width="50%"}
### Regression 📈
**Predicting numbers**
- House prices: $150K, $200K, $175K
- Sales volume: 1,000 units, 1,500 units  
- Temperature: 25°C, 30°C, 22°C
- Stock returns: +5%, -2%, +8%
:::

::: {.column width="50%"}
### Classification 🏷️
**Predicting categories**
- Email: Spam or Not Spam
- Loan decision: Approve or Reject  
- Customer: High/Medium/Low Value
- Medical diagnosis: Disease A, B, or Healthy
:::
:::

::: {.panel-tabset}

## R Code
```{r}
#| echo: true
#| eval: false
# Classification example: High wage prediction
Wage <- Wage %>%
  mutate(high_wage = ifelse(wage > 250, "High", "Low"))

# Logistic regression for classification
glm_fit <- glm(high_wage == "High" ~ age + education, 
               data = Wage, family = binomial)

# Get probabilities instead of just yes/no
pred_prob <- predict(glm_fit, type = "response")
```

## Python Code  
```python
from sklearn.linear_model import LogisticRegression
from sklearn.preprocessing import LabelEncoder

# Create binary outcome
wage_data['high_wage'] = (wage_data['wage'] > 250).astype(int)

# Prepare data
le = LabelEncoder() 
X = pd.DataFrame({
    'age': wage_data['age'],
    'education': le.fit_transform(wage_data['education'])
})
y = wage_data['high_wage']

# Fit classification model
clf = LogisticRegression()
clf.fit(X, y)

# Get probabilities
probabilities = clf.predict_proba(X)[:, 1]  # Prob of high wage
```

## Business Applications
**Classification is everywhere**:

- **Banking**: Loan approval (binary)
- **Marketing**: Customer value (low/medium/high)  
- **HR**: Resume screening (hire/reject)
- **E-commerce**: Product categories

**Key insight**: Classification gives you actionable categories, regression gives you precise numbers
:::

```{r}
#| echo: false
Wage <- Wage %>%
  mutate(highwage = ifelse(wage > 250, 1, 0))

glm_fit <- Wage %>%
  glm(highwage ~ age + education, data = ., family = binomial)

pred_data <- tibble(
  age = seq(20, 70, 1),
  education = "1. < HS Grad"
) %>%
  mutate(
    prob = predict(glm_fit, newdata = ., type = "response")
  )

pred_data %>%
  ggplot(aes(x = age, y = prob)) +
  geom_line(color = "blue", linewidth = 1.5) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "red") +
  annotate("text", x = 60, y = 0.52, label = "Decision Threshold (50%)", color = "red") +
  labs(
    title = "Probability of High Wage by Age",
    subtitle = "For workers with less than high school education",
    x = "Age",
    y = "P(High Wage)"
  ) +
  theme_minimal() +
  scale_y_continuous(labels = scales::percent_format()) +
  theme(
    plot.title = element_text(size = 13, face = "bold")
  )
```

---

## The Big Picture: Prediction vs. Interpretation Trade-off {.scrollable}

**Central tension in ML**: Accurate models are often hard to explain

::: {.columns}
::: {.column width="50%"}
### High Interpretability 🔍
**Simple models you can explain**
- Linear regression
- Decision trees (small ones)
- Simple rules

**When to use**:
- Regulatory requirements
- Medical decisions  
- Policy recommendations
- Building trust
:::

::: {.column width="50%"}
### High Prediction Accuracy 📈
**Complex "black box" models**
- Deep neural networks
- Random forests (large)
- Ensemble methods

**When to use**:
- Image recognition
- Recommendation systems
- High-stakes prediction
- When accuracy is paramount
:::
:::

**Real-world examples**:
- **Medical diagnosis**: Need interpretable models (doctor must explain)
- **Netflix recommendations**: Black box is fine (just works)
- **Credit scoring**: Regulated industry needs interpretable models
- **Fraud detection**: Accuracy matters more than explanation

---

## Key Takeaways for Economics & Business Students

::: {.incremental}
1. **Start with the business question**: Do you need prediction or understanding?

2. **Simple models first**: Linear regression beats fancy algorithms if you need interpretation

3. **Data quality matters more than algorithm choice**: Clean data + simple model > messy data + complex model

4. **Context is king**: A 90% accurate model is useless if you can't act on it

5. **Validate everything**: Your model is only as good as its performance on NEW data
:::

**Next steps**: Learn to evaluate model performance and avoid common pitfalls

---

## What's Coming Next

In future lectures, we'll cover:

::: {.incremental}
- **Model Assessment**: How do we know if our model is good?
- **Bias-Variance Trade-off**: Why simple models sometimes win
- **Cross-validation**: Testing models the right way  
- **Overfitting**: When models memorize instead of learn
- **Resampling Methods**: Making the most of limited data
:::

**Homework preview**: Apply these concepts to a real dataset from your country/region

---

## End

**Remember**: Machine learning is a tool, not magic. Success comes from: 

- Understanding your business problem 

- Choosing the right approach  

- Validating your results 

- Communicating findings clearly

*Questions?* 🤔
